---
title: "Plotting a phylogeny in R with APE library and various R base functions"
author: "Sebastien Renaut"
date: '2017-11-02'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,fig.width=7, fig.height=7)
```

```{r}
#There are many ways and options to plot phylogenies. Here I simply show one way, and show a few options to modify your phylogeny to highlight certain group of interest. 
```

```{r}
#library and dataset
library(ape)
library(phytools)
data(bird.orders)

#slightly better than default plot
phylo = plot.phylo(bird.orders, edge.width = 2, label.offset = 1)

###
###OPTION 1
###
#first find out where are the groups, by adding y-axis and looking at coordinates
phylo = plot.phylo(bird.orders, edge.width = 3, label.offset = 1)
axis(2,at =seq(from =0, to = phylo$y.lim[2]),cex.axis = 0.5)
axis(1,at =seq(from =0, to = phylo$x.lim[2]),cex.axis = 0.5)
xmin = phylo$x.lim[1]
xmax = phylo$x.lim[2] - 10 #so the names are not boxed

#add boxes based on the y axis coordinates, the x-axis coord. are based on the phylogeny, and your groups of interest
rect(xmin,13.5,xmax,23.5, col = "red",density = 30) #group 1
rect(xmin,0.5,xmax,5.5,col = "blue",density = 30) #group 2
rect(xmin,6.5,xmax,12.5,col = "purple",density = 30) #group 3

#it looks good, save it.
#dev.print(dev = pdf, "phylo_option1.pdf")
#dev.off()

###
###OPTION 2 (more complex set up, but better looking)
###
plot.new()
plot.window(xlim = phylo$x.lim,ylim = phylo$y.lim)

#three colors vector (red, blue, purple in hexadecimal, with last two digits as the transparency)
colors = c("#FF000060","#0000ff60","#551a8b60")

#add rectangles
rect(phylo$x.lim[1],0.5,phylo$x.lim[2],5.5,col = colors[1],border = F) #group 1
rect(phylo$x.lim[1],6.5,phylo$x.lim[2],12.5,col = colors[2], border = F) #group 2
rect(phylo$x.lim[1],13.5,phylo$x.lim[2],23.5, col = colors[3],border = F) #group 3

par(new = T)
plot.phylo(bird.orders, edge.width = 2, label.offset = 1)

#add a x axis that looks decent according to what we learned in the previous graph
axis(1,at =seq(from =0, to = xmax, by = 2),cex.axis = 0.5)

#it looks good, save it.
#dev.print(dev = pdf, "phylo_option2.pdf")
#dev.off()

###
###OPTION 3 (several plots combined in a layout)
###
### First save default parameters, for resetting if necessary later...
def.par <- par(no.readonly = TRUE) 

#setup layout as a 3 X 3 matrix with 3 regions (region 1 for the phylo tree, 2 for a x/y scatter and 3 for legend)
m = matrix(c(2,1,1,2,1,1,3,1,1), 3, 3, byrow = TRUE)

par(def.par)
layout(m)
#save plot.phylo parameters, without plotting
phylo = plot.phylo(bird.orders, edge.width = 2, label.offset = 1, plot = F, cex = 1)

#REGION 1#
#add boxes based on the y axis coordinates, the x-axis coord. are based on the phylogeny, and your groups of interest
rect(phylo$x.lim[1],0.5,phylo$x.lim[2],5.5,col = colors[1],border = F) #group 1
rect(phylo$x.lim[1],6.5,phylo$x.lim[2],12.5,col = colors[2], border = F) #group 2
rect(phylo$x.lim[1],13.5,phylo$x.lim[2],23.5, col = colors[3],border = F) #group 3

#add phylo on top
par(new = T)
phylo = plot.phylo(bird.orders, edge.width = 2, label.offset = 1, cex = 1)

#add a x axis that looks decent
axis(1,at =seq(from =0, to = xmax, by = 2),cex.axis = 0.5)

#Make the segment for Turniciformes dashed
#find out the x/y coordinates of the segment, color it white, then dashed red on top
segments(1,6,28,6,col = "white", lwd = 4, lty = 1)
segments(1,6,28,6,col = "black", lwd = 4, lty = 3)

#REGION 2#
#set up some wider margins for region # 2 of layout
par(mar = c(11,4,11,0))

#set up some fake data (say fake PCA data) for the 3 groups
pca = cbind(c(rnorm(6,mean=0,sd =1), rnorm(7,mean=2,sd =0.5),rnorm(10,mean=3,sd =0.5)),c(rnorm(6,mean=0,sd =1), rnorm(7,mean=0,sd =0.3),rnorm(10,mean=3,sd =0.2)))
rownames(pca) = bird.orders$tip.label

#colors scheme according to boxes in phylo tree (5 red,1 black, 6 blue, 1 black, 10 purple)
colors_according_to_tree = c(rep(colors[1],5),"black",rep(colors[2],6),"black",rep(colors[3],10))

#add another x/y scatter plot in region # 2 of layout
plot(pca,col = colors_according_to_tree,lwd =3,pch = 21, ylab = "random variable 1", xlab = "random variable 2")

#REGION 3#
#reset margins
par(mar = c(5,4,4,3))

#empty plot in region 3
plot(1, type="n", xlab="", ylab="", xlim=c(0, 1), ylim=c(0, 1),xaxt="n",yaxt= "n", bty = "n")

#add a legend here
legend(0.1,1.5,legend = c("group1","group2","group3"),fill = colors,xpd = T, cex = 2)

#it looks good, save it
#dev.print(dev = pdf, "phylo_option3.pdf")
#dev.off()

```



###add a trait value to the tree and plot it. 
trait = c(sample(c(50:100),2),c(sample(0:50,3)),sample(c(50:100),10),sample(c(0:50),8))
names(trait) = bird.orders$tip.label
tree.trait <- contMap(bird.orders, trait, plot = FALSE)
tree.trait<-setMap(tree.trait, colors=c("blue", "white", 'red'))
plot(tree.trait, type = "phylogram", legend = 0.4*max(nodeHeights(tree)), fsize = c(0.4, 0.7), lwd=3)

###add pie charts 
plot(obj, type = "phylogram", legend = 0.4*max(nodeHeights(tree)), fsize = c(0.4, 0.7), lwd=3)
tiplabels(pch=15, cex=.5, col=traits$habit_binary)

###add tip labels 
plot(obj, type = "phylogram", legend = 0.4*max(nodeHeights(tree)), fsize = c(0.4, 0.7), lwd=3)
tiplabels(pch=15, cex=.5, col=sample(c(1:3),120))

###add legends
legend(0.4,100,legend = c("habitat 1","habitat 2","habitat 3"),fill = c("red","blue","green"))
